Informing policy via dynamic models: Cholera in Haiti

### 2023-06-12 ms.Rnw
started revisions

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\theta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}
\newcommand\tHur{t_{hm}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{\bullet}
\newcommand\code[1]{\texttt{#1}}
\newcommand\paramVec{\theta}

@@

Per-capita transition rates are given in Equations~\ref{model1:SE}-\ref{model1:birth}:
\begin{eqnarray}
\label{model1:SE}
\mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t),
\\[-3pt]
\label{model1:EI}
\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\symptomFrac_{\vaccCounter}(t)\big),
\\[-3pt]
\label{model1:EA}
\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \symptomFrac_{\vaccCounter}(t),
\\[-3pt]
\label{model1:toR}
\mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR,
\\[-3pt]
\label{model1:RS}
\mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS,
\\[-3pt]
\label{model1:vacc}
\mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t),
\\[-3pt]
\label{model1:death}
\mu_{S_{\vaccCounter}\demography} &=& \mu_{E_{\vaccCounter}\demography} = \mu_{I_{\vaccCounter}\demography} = \mu_{A_{\vaccCounter}\demography}=\mu_{R_{\vaccCounter}\demography} = \delta,
\\
\label{model1:birth}
\mu_{\demography S_0} &=& \muBirth,
\end{eqnarray}
where $\vaccCounter\in \seq{0}{\vaccClass}$.
Here, $\mu_{AB}$ is a transition rate from compartment $A$ to $B$.
We have an additional demographic source and sink compartment $\demography$ modeling entry into the study population due to birth or immigration, and exit from the study population due to death or immigration.
Thus, $\mu_{A\demography}$ is a rate of exiting the study population from compartment $A$ and $\mu_{\demography B}$ is a rate of entering the study population into compartment $B$.

In Model~1, the advantage afforded to vaccinated individuals is an increased probability that an infection is asymptomatic.
Conditional on infection status, vaccinated individuals are also less infectious than their non-vaccinated counterparts by a rate of $\asymptomRelativeInfect = 0.05$ in Eq.~\eqref{model1:lambda}.
In \eqref{model1:EA} and~\eqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\symptomFrac_{\vaccCounter}(t) = c \, \theta^*(t-\tau_{\vaccCounter})$
where $\theta^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.

@@

Individuals move from department $u$ to $v$ at rate $T_{uv}$, and aquatic cholera moves at rate $T^W_{uv}$.
The nonzero transition rates are
\begin{eqnarray}
\label{model2:mu_SE}
\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda,
\\[-3pt]
\label{model2:mu_EI}
\mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI,
\\[-3pt]
\label{model2:mu_IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR,
\\[-3pt]
\label{model2:RS}
\mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS,
\\[-3pt]
\label{model2:transport}
\mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv},
\\[-3pt]
\label{model2:omega1}
\mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1,
\\[-3pt]
\label{model2:omega2}
\mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2,
\\[-3pt]
\label{model2:to_W}
\mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\},
\\[-3pt]
\label{model2:from_W}
\mu_{W_u\demography} &=& \Wremoval,
\\[-3pt]
\label{model2:water_transport}
\mu_{W_uW_{\varv}} &=& w_r T^W_{u\varv}.
\end{eqnarray}
In \eqref{model2:transport} the spatial coupling is specified by a gravity model,
\begin{equation}
\label{model2:gravity}
T_{u\varv} = v_{\mathrm{rate}} \times \frac{\mathrm{Pop}_u \mathrm{Pop}_{\varv}}{D_{u\varv}^2},
\end{equation}
where $\mathrm{Pop}_u$ is the mean population for department $u$,
$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$  was treated as a fixed constant.
In \eqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in \eqref{model2:lambda}.
The starting value for each element of the latent state vector $\bm{X}^{(2)}(0)$ are set to zero except for $I_{u0}(0) = y_u^*(0)/\reportRate$ and $R_{u0}(0) = \mathrm{Pop}_u - I_{u0}(0)$, where $y_u^*(0)$ is the reported number of cholera cases in department $u$ at time $t = 0$.
Reported cases are assumed to come from a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
See Sec.~\ref{sec:fit2} and Sec.~S3.2 of the supplement for more details.

@@

The latent state is described as $\bm{X}^{(3)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u\vaccCounter}(t),\allowbreak A_{u\vaccCounter}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
Here, $\vaccCounter=0$ corresponds to unvaccinated, $\vaccCounter=2j-1$ corresponds to a single dose on the $j$th vaccination campaign in unit $u$ and $\vaccCounter=2j$ corresponds to receiving two doses on the $j$th vaccination campaign.
$k\in\seq{1}{3}$ models non-exponential duration in the recovered class before waning of immunity.
The force of infection is
\begin{equation}
\label{eq:model3:foi}
\lambda_u(t) = \left(\Wbeta{_u} + 1_{(t \geq \tHur)}\Whur{_u} e^{-\hHur{u}\left(t - \tHur\right)}\right) \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv 0}(t)+ \asymptomRelativeInfect A_{\varv 0}(t)\big),
\end{equation}
where $\tHur$ is the time Hurricane Matthew struck Haiti \citep{ferreirai16}, and $1_{(A)}$ is the indicator function for event $A$.
In \cite{lee20}, $\Whur{_u}$ and $\hHur{u}$ were set to zero for all $u$;
the need to account for the effect Hurricane Matthew had on cholera transmission for this model is explored in Sec.~S5 of the supplement.

Per-capita transition rates are given in Equations~\ref{eq:model3:SI}--\ref{eq:model3:Decay}.
\begin{eqnarray}
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
\\[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u\vaccCounter}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\eta_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
\\[-3pt]
\label{eq:model3:IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter 1}} &=& \mu_{A_{u\vaccCounter}R_{u\vaccCounter 1}} = \muIR,
\\[-3pt]
\label{eq:model3:IS}
\mu_{I_{u\vaccCounter}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u\vaccCounter}S_{u0}} = \muDeath
\\[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
\\[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
\\[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u0}(t)+ \asymptomRelativeShed A_{u0}(t) \big],
\\[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval.
\end{eqnarray}
As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In \eqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
All deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.

@@

# First Search RW
reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
names(reg_rw_1.sd) <- est_param_names_expanded
chol_rw_1.sd <- do.call(rw.sd, reg_rw_1.sd)

# Second Search RW
reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
names(reg_rw_2.sd) <- est_param_names_expanded
chol_rw_2.sd <- do.call(rw.sd, reg_rw_2.sd)

# Third Search RW
reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
names(reg_rw_3.sd) <- est_param_names_expanded
chol_rw_3.sd <- do.call(rw.sd, reg_rw_3.sd)

@@

Recent information about a dynamic system should be more relevant for a forecast than older information.
This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \citep{king15}.
That may explain why \citet{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
$f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\theta)$
where $\hat\theta$ is a vector of calibrated parameters.
The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.

Uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20}.
Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
For frequentist inference, one possible approach is obtaining model forecasts from various values of $\theta$, where the values of $\theta$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \theta)$.
In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.

@@

  & {\fixed $\symptomFrac_z(t)=c\theta^*(t-\tau_d)$}

>>>>

%%%% parameters %%%%%%%%%%%
\newcommand\Wsat{W_{\mathrm{sat}}}
\newcommand\muIR{\mu_{IR}}
\newcommand\muEI{\mu_{EI}}
\newcommand\transmission{\beta}
\newcommand\seasAmplitude{a}
\newcommand\rainfallExponent{r}
\newcommand\muRS{\mu_{RS}}
\newcommand\vaccineEfficacy{\vartheta}
\newcommand\muBirth{\mu_S}
\newcommand\muDeath{\delta}
\newcommand\choleraDeath{\delta_{C}}
\newcommand\symptomFrac{f}
\newcommand\asymptomRelativeInfect{\epsilon}
\newcommand\asymptomRelativeShed{\epsilon_{W}}
\newcommand\Wbeta[1]{\beta_{W#1}}
\newcommand\Whur[1]{\beta_{W#1}^{hm}}
\newcommand\hHur[1]{h_{#1}^{hm}}
\newcommand\tHur{t_{hm}}
\newcommand\Wremoval{\delta_W}
\newcommand\Wshed{\mu_W}
\newcommand\mixExponent{\nu}
\newcommand\sigmaProc{\sigma_{\mathrm{proc}}}
\newcommand\reportRate{\rho}
\newcommand\obsOverdispersion{\psi}
\newcommand\phaseParm{\phi}
\newcommand\transmissionTrend{\zeta}
\newcommand\Binit{\xi}
\newcommand\vaccClass{Z}
\newcommand\vaccCounter{z}
\newcommand\modelCounter{m}
\newcommand\missing{---}
\newcommand\fixed{\color{blue}}
\newcommand\demography{\bullet}
\newcommand\code[1]{\texttt{#1}}
\newcommand\paramVec{\theta}
\newcommand\childReduce{q}

@@

Per-capita transition rates are given in Equations~\ref{model1:SE}-\ref{model1:birth}:
\begin{eqnarray}
\label{model1:SE}
\mu_{S_{\vaccCounter}E_{\vaccCounter}} &=& \lambda(t),
\\[-3pt]
\label{model1:EI}
\mu_{E_{\vaccCounter}I_{\vaccCounter}} &=& \muEI\big(1-\vaccineEfficacy_\vaccCounter(t)\big),
\\[-3pt]
\label{model1:EA}
\mu_{E_{\vaccCounter}A_{\vaccCounter}} &=& \muEI\, \vaccineEfficacy_\vaccCounter(t),
\\[-3pt]
\label{model1:toR}
\mu_{I_{\vaccCounter}R_{\vaccCounter}} &=& \mu_{A_{\vaccCounter}R_{\vaccCounter}} = \muIR,
\\[-3pt]
\label{model1:RS}
\mu_{R_{\vaccCounter}S_{\vaccCounter}} &=& \muRS,
\\[-3pt]
\label{model1:vacc}
\mu_{S_0S_{\vaccCounter}} &=& \mu_{E_0E_{\vaccCounter}} = \mu_{I_0I_{\vaccCounter}} = \mu_{A_0A_{\vaccCounter}} = \mu_{R_0R_{\vaccCounter}} = \eta_{\vaccCounter}(t),
\\[-3pt]
\label{model1:death}
\mu_{S_{\vaccCounter}\demography} &=& \mu_{E_{\vaccCounter}\demography} = \mu_{I_{\vaccCounter}\demography} = \mu_{A_{\vaccCounter}\demography}=\mu_{R_{\vaccCounter}\demography} = \delta,
\\
\label{model1:birth}
\mu_{\demography S_0} &=& \muBirth,
\end{eqnarray}
where $\vaccCounter\in \seq{0}{\vaccClass}$.
Here, $\mu_{AB}$ is a transition rate from compartment $A$ to $B$.
We have an additional demographic source and sink compartment $\demography$ modeling entry into the study population due to birth or immigration, and exit from the study population due to death or immigration.
Thus, $\mu_{A\demography}$ is a rate of exiting the study population from compartment $A$ and $\mu_{\demography B}$ is a rate of entering the study population into compartment $B$.

In Model~1, the advantage afforded to vaccinated individuals is an increased probability that an infection is asymptomatic.
Conditional on infection status, vaccinated individuals are also less infectious than their non-vaccinated counterparts by a rate of $\asymptomRelativeInfect = 0.05$ in Eq.~\eqref{model1:lambda}.
In \eqref{model1:EA} and~\eqref{model1:EI} the asymptomatic ratio for non-vaccinated individuals is set $\symptomFrac_0(t)=0$, so that the asymptomatic route is reserved for vaccinated individuals.
For $\vaccCounter\in\seq{1}{\vaccClass}$, the vaccination cohort $\vaccCounter$ is assigned a time $\tau_{\vaccCounter}$, and we take $\vaccineEfficacy_\vaccCounter(t) = c \, \vaccineEfficacy^*(t-\tau_{\vaccCounter})$
where $\vaccineEfficacy^*(t)$ is efficacy at time $t$ since vaccination for adults, taken from \cite{lee20}, Table~S4, and $c=\big(1-(1-0.4688)\times 0.11\big)$ is a correction to allow for reduced efficacy in the 11\% of the population aged under 5 years.
Single and double vaccine doses were modeled by changing the waning of protection; protection was assumed to be equal between single and double dose until 52 weeks after vaccination, at which point the single dose becomes ineffective.

@@

Individuals move from department $u$ to $v$ at rate $T_{uv}$, and aquatic cholera moves at rate $T^W_{uv}$.
The nonzero transition rates are
\begin{eqnarray}
\label{model2:mu_SE}
\mu_{S_{u\vaccCounter}E_{u\vaccCounter}} &=& \vaccineEfficacy_\vaccCounter \, \lambda,
\\[-3pt]
\label{model2:mu_EI}
\mu_{E_{u\vaccCounter}I_{u\vaccCounter}} &=& \symptomFrac\muEI, \quad \mu_{E_{u\vaccCounter}A_{u\vaccCounter}} = (1-\symptomFrac)\muEI,
\\[-3pt]
\label{model2:mu_IR}
\mu_{I_{u\vaccCounter}R_{u\vaccCounter}} &=& \mu_{A_{u\vaccCounter}R^A_{u\vaccCounter}} = \muIR,
\\[-3pt]
\label{model2:RS}
\mu_{R_{u\vaccCounter}S_{u\vaccCounter}} &=& \mu_{R^A_{u\vaccCounter}S_{u\vaccCounter}} = \muRS,
\\[-3pt]
\label{model2:transport}
\mu_{S_{u\vaccCounter}S_{{\varv}\vaccCounter}} &=& \mu_{E_{u\vaccCounter}E_{{\varv}\vaccCounter}} = \mu_{I_{u\vaccCounter} I_{{\varv}\vaccCounter}} = \mu_{A_{u\vaccCounter}A_{{\varv}\vaccCounter}} = \mu_{R_{u\vaccCounter}R_{{\varv}\vaccCounter}} = \mu_{R^A_{u\vaccCounter} R^A_{{\varv}\vaccCounter}} = T_{u\varv},
\\[-3pt]
\label{model2:omega1}
\mu_{S_{u1}S_{u0}} &=& \mu_{S_{u3}S_{u0}} = \omega_1,
\\[-3pt]
\label{model2:omega2}
\mu_{S_{u2}S_{u0}} &=& \mu_{S_{u4}S_{u0}} = \omega_2,
\\[-3pt]
\label{model2:to_W}
\mu_{\demography W_u} &=& \Wshed \left\{ \sum_{\vaccCounter=0}^4 I_{u\vaccCounter}(t) + \asymptomRelativeShed \sum_{\vaccCounter=0}^4 A_{u\vaccCounter}(t) \right\},
\\[-3pt]
\label{model2:from_W}
\mu_{W_u\demography} &=& \Wremoval,
\\[-3pt]
\label{model2:water_transport}
\mu_{W_uW_{\varv}} &=& w_r T^W_{u\varv}.
\end{eqnarray}
In \eqref{model2:transport} the spatial coupling is specified by a gravity model,
\begin{equation}
\label{model2:gravity}
T_{u\varv} = v_{\mathrm{rate}} \times \frac{\mathrm{Pop}_u \mathrm{Pop}_{\varv}}{D_{u\varv}^2},
\end{equation}
where $\mathrm{Pop}_u$ is the mean population for department $u$,
$D_{u\varv}$ is a distance measure estimating average road distance between randomly chosen members of each population, and $v_{\mathrm{rate}}= 10^{-12} \,\mbox{km$^2$yr$^{-1}$}$  was treated as a fixed constant.
In \eqref{model2:water_transport}, $T^W_{u\varv}$ is a measure of river flow between departments.
The unit of $W_u(t)$ is cells per ml, with dose response modeled via a saturation constant of $\Wsat$ in \eqref{model2:lambda}.
In \eqref{model2:mu_SE}, $\vaccineEfficacy_\vaccCounter$ denotes the vaccine efficacy for each vaccination campaign $\vaccCounter \in \vaccClass$, with $\vaccineEfficacy_0 = 1$, $\vaccineEfficacy_1 = 1 - 0.429 \childReduce$, $\vaccineEfficacy_2 = 1 - 0.519 \childReduce$, $\vaccineEfficacy_3 = 1 - 0.429$, and $\vaccineEfficacy_4 = 1 - 0.519$
Here, $\childReduce = 0.4688$ represents the reduced efficacy of the vaccination for children under the age of five years, and the values $0.429$ and $0.519$ are the median effectiveness of one and two doses over their effective period respectively, according to Table S4 in the supplement material of \citet{lee20}.
Because vaccine efficacy remains constant, individuals in this model transition from a vaccinated compartment to the susceptible compartment at the end of the vaccine coverage period.

The starting value for each element of the latent state vector $\bm{X}^{(2)}(0)$ are set to zero except for $I_{u0}(0) = y_u^*(0)/\reportRate$ and $R_{u0}(0) = \mathrm{Pop}_u - I_{u0}(0)$, where $y_u^*(0)$ is the reported number of cholera cases in department $u$ at time $t = 0$.
Reported cases are assumed to come from a log-normal distribution, with the log-scale mean equal to the reporting rate $\reportRate$ times the number of newly infected individuals.
See Sec.~\ref{sec:fit2} and Sec.~S3.2 of the supplement for more details.

@@

The latent state is described as $\bm{X}^{(3)}(t) = \big(S_{u\vaccCounter}(t),\allowbreak I_{u}(t),\allowbreak A_{u}(t),\allowbreak R_{u\vaccCounter k}(t),\allowbreak W_u(t),\allowbreak u \in \seq{0}{U},\allowbreak \vaccCounter \in \seq{0}{4},\allowbreak k \in \seq{1}{3}\big)$.
Here, $\vaccCounter=0$ corresponds to unvaccinated, $\vaccCounter=2j-1$ corresponds to a single dose on the $j$th vaccination campaign in unit $u$ and $\vaccCounter=2j$ corresponds to receiving two doses on the $j$th vaccination campaign.
$k\in\seq{1}{3}$ models non-exponential duration in the recovered class before waning of immunity.
The force of infection is
\begin{equation}
\label{eq:model3:foi}
\lambda_u(t) = \left(\Wbeta{_u} + 1_{(t \geq \tHur)}\Whur{_u} e^{-\hHur{u}\left(t - \tHur\right)}\right) \frac{W_u(t)}{1+W_u(t)} + \transmission_u \sum_{{\varv}\neq u}\big(I_{\varv}(t)+ \asymptomRelativeInfect A_{\varv}(t)\big),
\end{equation}
where $\tHur$ is the time Hurricane Matthew struck Haiti \citep{ferreirai16}, and $1_{(A)}$ is the indicator function for event $A$.
In \cite{lee20}, $\Whur{_u}$ and $\hHur{u}$ were set to zero for all $u$;
the need to account for the effect Hurricane Matthew had on cholera transmission for this model is explored in Sec.~S5 of the supplement.

Per-capita transition rates are given in Equations~\ref{eq:model3:SI}--\ref{eq:model3:Decay}.
\begin{eqnarray}
\label{eq:model3:SI}
\mu_{S_{u\vaccCounter}I_{u}} &=& \symptomFrac \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \, d\Gamma/dt,
\\[-3pt]
\label{eq:model3:SA}
\mu_{S_{u\vaccCounter}A_{u}} &=& (1-\symptomFrac) \,  \lambda_u \big(1-\vaccineEfficacy_{u\vaccCounter}(t)\big) \,  d\Gamma/dt,
\\[-3pt]
\label{eq:model3:IR}
\mu_{I_{u}R_{u\vaccCounter 1}} &=& \mu_{A_{u}R_{u\vaccCounter 1}} = \muIR,
\\[-3pt]
\label{eq:model3:IS}
\mu_{I_{u}S_{u0}} &=& \muDeath + \choleraDeath, \hspace{2mm} \mu_{A_{u}S_{u0}} = \muDeath
\\[-3pt]
\label{eq:model3:RRnext}
\mu_{R_{u\vaccCounter 1}R_{u\vaccCounter 2}} &=& \mu_{R_{u\vaccCounter 2}R_{u\vaccCounter 3}} = 3\muRS,
\\[-3pt]
\label{eq:model3:RS}
\mu_{R_{u\vaccCounter k}S_{u0}} &=& \muDeath + 3\muRS \, \mathbf{1}_{\{k=3\}},
\\[-3pt]
\label{eq:model3:water}
\mu_{{\demography}W_u} &=& \big[1 + \seasAmplitude \big(J_u(t))^r \big] \mathrm{Den}_u \, \Wshed \big[ I_{u}(t)+ \asymptomRelativeShed A_{u}(t) \big],
\\[-3pt]
\label{eq:model3:Decay}
\mu_{W_u\demography} &=& \Wremoval.
\end{eqnarray}
As with Model~1, $d\Gamma_u(t)/dt$ is multiplicative Gamma-distributed white noise in \myeqref{eq:model3:SI} and \myeqref{eq:model3:SA}.
In \eqref{eq:model3:water}, $J_u(t)$ is a dimensionless measurement of precipitation that has been standardized by dividing the observed rainfall at time $t$ by the maximum recorded rainfall in department $u$ during the epidemic, and $\mathrm{Den}_u$ is the population density.
Demographic stochasticity is accounted for by modeling non-cholera related death rate $\muDeath$ in each compartment, along with an additional death rate $\choleraDeath$ in \myeqref{eq:model3:IS} to account for cholera induced deaths among infected individuals.
All deaths are balanced by births into the susceptible compartment in \myeqref{eq:model3:IS} and \myeqref{eq:model3:RS}, thereby maintaining constant population in each department.

Similar to Model~1, there are no distinct compartments for individuals under five years of age, and the vaccination efficacy is taken as a age adjusted weighted average of the efficacy for individuals both over and under five years of age:  $\vaccineEfficacy_{u \vaccCounter}(t) = c\vaccineEfficacy^*(t - \tau_{uz})$, where $\tau_{uz}$ is the vaccination time for unit $u$ and vaccination campaign $z$.
The value $c$ and the function $\vaccineEfficacy^*$ are equivalent to those described in Sec.~\ref{sec:model1}.

@@

# First Search RW
reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
names(reg_rw_1.sd) <- est_param_names_expanded
chol_rw_1.sd <- do.call(rw_sd, reg_rw_1.sd)

# Second Search RW
reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
names(reg_rw_2.sd) <- est_param_names_expanded
chol_rw_2.sd <- do.call(rw_sd, reg_rw_2.sd)

# Third Search RW
reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
names(reg_rw_3.sd) <- est_param_names_expanded
chol_rw_3.sd <- do.call(rw_sd, reg_rw_3.sd)

@@

Recent information about a dynamic system should be more relevant for a forecast than older information.
This assertion may seem self-evident, but it is not the case for deterministic models, for which the initial conditions together with the parameters are sufficient for forecasting, and so recent data do not have special importance.
Epidemiological forecasts based on deterministic models are not uncommon in practice, despite their limitations \citep{king15}.
That may explain why \citet{lee20} chose to obtain forecasts by simulating the calibrated models forward from initial conditions.
Here, we compare with a forecast projected from the filtering distribution: to obtain a forecast using POMP model $m$ at a collection of times $t_{N+1:N+s}$, where $N$ is the index for the last available data, we simulate forward from a draw from
$f^{(m)}_{\bm{X}_N|\bm{Y}_{1:N}}(\bm{x}_{N} | \bm{y}^*_{1:N} ; \hat\paramVec)$
where $\hat\paramVec$ is a vector of calibrated parameters.
The decision not to do this partially explains the unsuccessful forecasts of \cite{lee20}: their Table~S7 shows that the subset of their simulations which were consistent with observing zero cases in 2019 also accurately predicted the prolonged absence of detected cholera.

Uncertainty in just a single parameter can lead to drastically different forecasts \citep{saltelli20}.
Therefore, parameter uncertainty should also be considered when obtaining model forecasts to influence policy.
If a Bayesian technique is used for parameter estimation, a natural way to account for parameter uncertainty is to obtain simulations from the model where each simulation is obtained using parameters drawn from the estimated posterior distribution.
For frequentist inference, one possible approach is obtaining model forecasts from various values of $\paramVec$, where the values of $\paramVec$ are sampled proportionally according to their corresponding likelihoods \citep{king15}; see Sec.~S7 of the supplement material for more details.
Both of these approaches share the similarity that parameters are chosen for the forecast approximately in proportion to their corresponding value of the likelihood function, $f^{(m)}_{\bm{Y}_{1:N}}(\bm{y}_{1:N}^*; \paramVec)$.
In this analysis, we do not construct forecasts accounting for parameter uncertainty as our focus is on the estimation and diagnosis of mechanistic models.
Furthermore, we use the projections from a single point estimate to highlight the deficiency of deterministic models that the only variability in model projections is a result of parameter uncertainty, which can lead to over-confidence in forecasts \cite{king15}.

@@

  & {\fixed $\symptomFrac_z(t)=c\vaccineEfficacy^*(t-\tau_d)$}
